Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals

### 2018-09-26 paper/paper.tex
Surrogate model used for PE runs. Many paper edits. References now complete.

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors has demonstrated that tidal interactions during the inspiral can be measured with enough precision to constrain the radius and equation of state~\cite{DeEtA, PEPaper, EOSPaperl} of NSs. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of the radius and EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}. \red{[[Mention A+ being funded? How will this improve each parameter?]]}

@@

These results are obtained using a Bayesian analysis that compares the data to parameterized waveform models. 
Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.

@@

In addition, to avoid systematic errors that can be as large as statistical errors, highly accurate models are needed that include all relevant physical effects, including point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. For GW170817, a set of four frequency-domain analytic or semi-analytic waveform models were used as templates (see Table I of~\cite{PEPaper}) for the main results. These models are based on spinning binary black hole (BBH) waveform models. They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2~\cite{}), the aligned-spin effective one body (EOB) formalism (SEOBNRT~\cite{}) or the phenomenological formalism for aligned-spin (PhenomDNRT~\cite{}) and precessing spin (PhenomPNRT~\cite{}). Tidal interactions, on the other hand, were included simply by adding a correction to the phase of each waveform. For the PN TaylorF2 waveform the analytic 5PN and 6PN order terms were added. For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in~\cite{PEPaper}) was used for the phase correction~\cite{DietrichBernuzziTichy2017, NewNRTidalPaper}.

@@

In this paper we will use a new implementation of the effective one body (EOB) formalism, that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. 
%Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model. 
%Recently two versions have become available that include both tidal effects and aligned spin and are implemented under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}. 
The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note an alternative model, \texttt{TEOBResumS}~\cite{Nagar2017, NewPaper2018}, that uses a slightly different point-particle and spin prescription. For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger. \red{[[Must mention evaluation time to segue into next 2 paragraphs.]]}

@@

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. For GW170817~\cite{PEPaper}, an alternative highly parallelized code, \texttt{RapidPE}, was used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{RapidPE, RIFT}. Unfortunately, while the results were broadly consistent with the MCMC results when using the same waveform models, there were noticeable differences. In particular the shape of the PDF was different, and the credible intervals for the tidal parameters were shifted downward by $\sim 10$\%~\cite{PEPaper}. Another limitation is that RapidPE does not currently marginalize over calibration errors. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. In addition, different methods of formulating the problem can allow for faster integration~\cite{RushTheInspiral}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T}, discussed below, that bypasses the need for specialized parameter estimation tools and optimization techniques.

@@

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
% whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. 
The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}
%method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. 
In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

The final step in building a surrogate is to interpolate between waveform parameters $\bx$. Previous work focused on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the EOB waveform here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible 
%\MP{I think that this is too strong; Blackman17a also perform 5D TP interpolation which contradicts this.}. 
Recent work on optimally choosing waveforms for analytic~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on a technique known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}. GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.

This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that does not require an online Fourier transform for each waveform evaluation.
%\MP{Is it really true that this is a significant speedup? FFT may still be cheaper than surrogate waveform generation.} 
It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}. The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

We organize the paper as follows. In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section IV. In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss applications and future improvements in Section V.

@@

More precisely, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}. This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, but is only significant for systems with large spin. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2017}.

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. If either component of the binary is a black hole, then we set the tidal polarizabilities to zero. Let $M_{A,B}$ be the masses bodies $A$ and $B$, and $M = M_A + M_B$ be the total mass of the binary. The tidal contribution, including quadrupolar and octupolar dynamical tides, is
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
\end{align}
\end{widetext}
Here, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$. Here, $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$. This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

@@

We now look for a suitable time $t_{\rm match}$ to stop the numerical integration and match with a post-merger waveform. In the point-mass model \texttt{SEOBNRv4}, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks, and $t_{\rm match}$ is chosen to be the time of the peak amplitude $t_{\rm peak}^{\rm amp}$. Here, we do not use nonquasicircular corrections tuned to BNS simulations \red{[[Ambiguous. Do we use them or not. Find out.]]}, so we have less control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral. We therefore choose the following definition for $t_{\rm match}$. Let $t_{\rm peak}^{\rm amp}$ be the earliest time when the amplitude $|h_{22}^{\textrm{insp-plunge}}|$ of the (2, 2) mode peaks, and let $t_{\rm peak}^{\rm freq}$ be the time when the frequency $\omega_{22}^{\rm insp-plunge}$ of the (2, 2) mode peaks. Then,
\begin{equation}
t_{\textrm{match}} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.\label{tmatch}
\end{equation}
For some waveforms, the amplitude does not peak. In this case, we choose $t_{\rm peak}^{\rm amp}$ to be the earliest time when the slope of the amplitude $\partial_t |h_{22}^{\rm insp-plunge}|$ reaches a minimum after having reached a peak.

@@

For the amplitude, we smoothly extend the waveform after $t_{\rm match}$ with a linear fit, and then taper the resulting amplitude. The linear extension is defined by
\begin{align}
\hat{A}(t)=\left\{\begin{array}{ll}
|h_{22}^{\rm insp-plunge}(t)|, & t \leq t_{\rm match}\,, \\
a + b (t - t_{\rm match}), & t > t_{\rm match}\,,
\end{array}\right.
\end{align}
where $a = |h_{22}^{\rm insp-plunge}(t_{\rm match})|$ and $b=\partial_t |h_{22}^{\rm insp-plunge}(t_{\rm match})|$. The tapering function is centered $15M$ after $t_{\rm match}$ and has a decay time $\tau = \pi/(2\omega_{\rm match})$ (where $\omega_{\rm match} = \omega_{22}^{\rm insp-plunge}(t_{\rm match})$) of one gravitational wave period. \red{[[As written, this is a decay time of 1/4 of a period. Figure out what is correct.]]} It is given by
\begin{equation}
W(t) = \frac{1}{1+\exp{[(t- t^{22}_{\textrm{match}}-15M)/\tau]}}\,.
\end{equation}
The final amplitude after windowing is then given by $|h_{22}(t)| = \hat A(t) W(t)$.

>>>>

\red{only use texttt for actual programs and functions. TaylorF2 should usually be standard text}

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} by the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors has demonstrated that tidal interactions during the inspiral can be measured with enough precision to constrain the radius and equation of state~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, BNSPE, EOSPaper} of NSs. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of the radius and EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}. 
%\red{[[Mention A+ being funded? How will this improve each parameter?]]}

@@

These results are obtained using Bayesian analyses that compare the data to parameterized waveform models. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. Most time-domain waveform implementations involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. Depending on the model, this can take seconds to hours for long BNS waveforms, so analytic frequency-domain approximations are often used instead.

@@

In addition, to avoid systematic errors that can be as large as statistical errors, highly accurate waveform models are needed that include all relevant physical effects, including point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. For GW170817, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either the aligned-spin post-Newtonian (PN) formalism (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions, on the other hand, were included simply by adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was used for the phase correction~\cite{DietrichBernuzziTichy2017}.

@@

In this paper we will use a new implementation of the effective one body (EOB) formalism, that treats matter effects consistently with the other parameters. It includes several dynamical effects beyond the standard adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. The model, named \texttt{SEOBNRv4T}, includes the tidally induced $\ell=2$ and 3 multipole moments, the induced $\ell=2$ and 3 $f$-mode resonances~\cite{Hinderer:2016eia,Steinhoff:2016rfi}, and the spin-induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2018}. This model agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. It also includes a prescription for tapering the end of the waveform after merger in a manner consistent with numerical simulations. We also note an alternative model, \texttt{TEOBResumS}~\cite{NagarBernuzziDelpozzo2018}, that uses a slightly different point-particle and spin prescription. For matter effects, the model includes the $\ell = 2, 3$, and 4 tidal terms and uses a method for re-summing these expressions. It also includes the spin induced-quadrupole term, but does not include the effect of dynamical tides from the induced $f$-mode resonances or a tapering to zero amplitude after merger. 
%\red{[[Must mention evaluation time to segue into next 2 paragraphs.]]}

@@

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. For GW170817~\cite{BNSPE}, an alternative highly parallelized code, \texttt{RapidPE}, was used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. Unfortunately, while the results were broadly consistent with the MCMC results when using the same waveform models, there were noticeable differences. In particular the shape of the PDF was different, and the credible intervals for the tidal parameters were shifted downward by $\sim 10$\%~\cite{BNSPE}. Another limitation is that RapidPE does not currently marginalize over calibration errors. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. In addition, different methods of formulating the problem can allow for faster integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

@@

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for rapidly evaluating waveform models. The essential idea is to construct a fitting function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the space of all waveforms using a reduced basis of orthonormal functions.
% whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. 
The two most common approaches for selecting a reduced basis have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}.
%method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. 
In this work, instead of using a reduced basis, we will instead use the fact that the Fourier transformed EOB waveform is approximately known analytically via the fast, analytic TaylorF2 model. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we can efficiently represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

The final step in building a surrogate is to interpolate between waveform parameters $\bx$. Previous work focused on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality: for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the EOB waveform here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible. Recent work on optimally choosing waveforms for analytic~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on a technique known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of nonspinning, eccentric BBH mergers~\cite{HuertaMooreKumar2017}. GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.

This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}. 
%The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

We organize the paper as follows. In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section IV. In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss applications and future improvements in Section V.

@@

For the model used in this paper, \texttt{SEOBNRv4T}, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. \red{It includes spin-orbit interactions up to 3PN order and spin-spin interactions up to 2PN order. In addition, for the \texttt{SEOBNRv4T} model, the spin-induced quadrupole moment~\cite{Poisson1998} is also included. This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, and is significant for systems with large spin.} \red{\sout{The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}. This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, but is only significant for systems with large spin. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2018}}}.

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. If either component of the binary is a black hole, then we set the tidal polarizabilities to zero. The tidal contribution, including quadrupolar and octupolar dynamical tides, is
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
\end{align}
\end{widetext}
Here, $M_{A,B}$ are the masses of bodies $A$ and $B$, $M = M_A + M_B$ is the total mass, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$. Here, $\hat{k}_{\ell, \textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$. This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

@@

We now look for a suitable time $t_{\rm match}$ to stop the numerical integration and match with a post-merger waveform. In the point-mass model, \texttt{SEOBNRv4}, nonquasicircular corrections to the inspiral-plunge signal guarantee that the waveform peaks, and $t_{\rm match}$ is chosen to be the time of the peak amplitude $t_{\rm peak}^{\rm amp}$. Here, we do not use nonquasicircular corrections tuned to BNS simulations, so we have less control over the behavior of Eq.~(\ref{hlminspplunge}) in the late inspiral. We therefore choose the following definition for $t_{\rm match}$. Let $t_{\rm peak}^{\rm amp}$ be the earliest time when the amplitude $|h_{22}^{\textrm{insp-plunge}}|$ of the (2, 2) mode peaks, and let $t_{\rm peak}^{\rm freq}$ be the time when the frequency $\omega_{22}^{\rm insp-plunge}$ of the (2, 2) mode peaks. Then,
\begin{equation}
t_{\textrm{match}} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.\label{tmatch}
\end{equation}
For some waveforms, the amplitude does not peak. In this case, we choose $t_{\rm peak}^{\rm amp}$ to be the earliest time when the slope of the amplitude $\partial_t |h_{22}^{\rm insp-plunge}|$ reaches a minimum after having reached a peak.

@@

For the amplitude, we smoothly extend the waveform after $t_{\rm match}$ with a linear fit, and then taper the resulting amplitude. The linear extension is defined by
\begin{align}
\hat{A}(t)=\left\{\begin{array}{ll}
|h_{22}^{\rm insp-plunge}(t)|, & t \leq t_{\rm match}\,, \\
a + b (t - t_{\rm match}), & t > t_{\rm match}\,,
\end{array}\right.
\end{align}
where $a = |h_{22}^{\rm insp-plunge}(t_{\rm match})|$ and $b=\partial_t |h_{22}^{\rm insp-plunge}(t_{\rm match})|$. The tapering function is centered $15M$ after $t_{\rm match}$ and has a decay time $\tau = 2\pi / \omega_{\rm match}$ (where $\omega_{\rm match} = \omega_{22}^{\rm insp-plunge}(t_{\rm match})$) of one gravitational wave period. It is given by
\begin{equation}
W(t) = \frac{1}{1+\exp{[(t- t^{22}_{\textrm{match}}-15M)/\tau]}}\,.
\end{equation}
The final amplitude after windowing is then given by $|h_{22}(t)| = \hat A(t) W(t)$.

For the phase, we smoothly extend the waveform frequency such that it agrees with the inspiral frequency at a time $t_{\rm freq} = t_{\rm match} - 20M$ before the matching time $t_{\rm match}$, but then stretches out the frequency evolution such that it only approaches $\omega_{\rm match}$ asymptotically. We define this frequency evolution as
\begin{align}
\omega(t) = \omega_{\rm match} - \Delta\omega \exp\left[ -(t - t_{\rm freq}) / (20M) \right],
\end{align}
where $\Delta\omega = \omega_{\rm match} - \omega_{\rm freq}$. Integrating the frequency, and requiring continuity at $t_{\rm freq}$, results in the final expression for the phase
\begin{align}
\phi_{22}(t) = \left\{\begin{array}{ll}
\phi_{22}^{\textrm{insp-plunge}}(t), & t \leq t_{\rm freq} \,, \\
\phi_{\rm freq} + \omega_{\rm match}(t - t_{\rm freq}) + 20M \Delta\omega \left\{ \right. & \\
  \left. \exp\left[ -(t - t_{\rm freq}) / (20M) \right] - 1\right\} \,,& t > t_{\rm freq} \, .
\end{array}\right.
\end{align}
